\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-03-01 main.Rmd
major changes of abstract after talking to Di

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data refers to measurements taken across space over time. In practice, spatial and temporal data many times need to be combined in order to form a spatio-temporal data set. However, these two types of data are collected at different observational units and this poses a problem on how to structure the data for different wrangling tasks. Analysts can either organise the data in a single table format that combines all the variables, or in two separate tables, each focusing on the spatial or temporal aspect of the data. Both formats are useful in exploratory data analysis, visualisation, and, modelling, however, it can be labourious to constantly switch between the two formats during the analysis and write code to cast the data into the desirable shape. This paper presents a new \proglang{R} package, \pkg{cubble}, that provides a new data structure to ease the conversion between the two formats in spatio-temporal data. The new data structure organises spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. \pkg{Cubble} also provides the capability for handling data with hierarchical structure, matching data from multiple sources, constructing interactive graphics, and performing spatio-temporal transformation. The paper will demonstrate \pkg{cubble} with examples using Australian climate weather stations, river level data, and climate reanalysis (ERA5) data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data refer to measurements taken across space and time. In practice, spatio-temporal data can be decomposed into a spatial and temporal component: at one time, we would select a spatial location and inspect the temporal trend; at other time, we might select one or multiple time value(s) and explore the spatial distribution. Ideally, we could make multiple maps and multiple time series to explore these together, however, doing all of these actions is complicated when data arrive fragmented in multiple objects. To make it easy to do all these tasks, ideally spatial and temporal variables are in a single data object that we can slice and dice in different ways to conduct different visualisations. This work suggests a new data structure, \code{cubble}, that uses a nested form and a long form to organise spatial and temporal variables in a single data object. The new data structure ensures that data in the two forms are synchronised while been flexible to explore the two aspect of spatio-temporal data. It also provides capability for handling data with hierarchical structure, matching data from multiple sources, constructing interactive graphics, and performing spatio-temporal transformation. The paper will demonstrate \pkg{cubble} with examples using Australian climate weather stations, river level data, and climate reanalysis (ERA5) data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

### 2022-03-01 main.tex
major changes of abstract after talking to Di

\Abstract{
Spatio-temporal data refers to measurements taken across space over
time. In practice, spatial and temporal data many times need to be
combined in order to form a spatio-temporal data set. However, these two
types of data are collected at different observational units and this
poses a problem on how to structure the data for different wrangling
tasks. Analysts can either organise the data in a single table format
that combines all the variables, or in two separate tables, each
focusing on the spatial or temporal aspect of the data. Both formats are
useful in exploratory data analysis, visualisation, and, modelling,
however, it can be labourious to constantly switch between the two
formats during the analysis and write code to cast the data into the
desirable shape. This paper presents a new \proglang{R} package,
\pkg{cubble}, that provides a new data structure to ease the conversion
between the two formats in spatio-temporal data. The new data structure
organises spatial and temporal variables as two forms of a single data
object so that they can be wrangled separately while synchronised as a
whole piece. \pkg{Cubble} also provides the capability for handling data
with hierarchical structure, matching data from multiple sources,
constructing interactive graphics, and performing spatio-temporal
transformation. The paper will demonstrate \pkg{cubble} with examples
using Australian climate weather stations, river level data, and climate
reanalysis (ERA5) data.
}

@@

      Nicolas Langrené\\
    BNU-HKBU United International College\\
    2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China\\
  E-mail: \href{mailto:nicolas@langrene.com}{\nolinkurl{nicolas@langrene.com}}\\

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time intervals. This type of data can naturally be
considered as a data cube with three axes: spatial identifier, temporal
identifier, and variable. In this cubic formation, space only occupies
one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

\caption[An illustration of different spatio-temporal data layout]{An illustration of different spatio-temporal data layout: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

@@

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} \citep{spacetime} is an attempt for such a
data object, but the spatial and temporal class it built on have now
been gradually replaced by more recent implementations. The \pkg{stars}
\citep{stars} package uses an array to wrangle spatio-temporal data, but
the dense array it uses can quickly exhaust the memory for data in the
irregular spatial grid. While there has not yet been a data object that
makes it easy to wrangle spatio-temporal data, there is a strong need
for such tools given the ubiquitous spatial and temporal data collected
in the society, i.e., climate observations from weather stations, air
quality variables from air monitoring stations. The demand for such
tools can also come from spatial data analysts wishing to incorporate
temporal data in their analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to
overcome the problem with organising variables in different levels when
working with spatio-temporal data. In the package a new data structure,
also called \code{cubble}, is proposed to organise spatial and temporal
variables as two forms of a single data object so that they can be
wrangled separately while synchronised as a whole piece. With the new
data object, analysts can spend less time writing codes to organise
spatial and temporal variables and focus more on the exploratory data
analysis itself. The software is available from the Comprehensive R
Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the main
design and functionality of \pkg{cubble}. Section 3 explains how cubble
deals with more advanced considerations including data with hierarchical
structure, data matching, how cubble fits with existing static and
interactive visualisation tools, and spatio-temporal data
transformation. Section 4 uses Australia weather station data and river
level data as examples to demonstrate the use with \pkg{cubble}. An
example on how \pkg{cubble} handles NetCDF data is also provided.
Section 5 concludes the paper.

>>>>

\Abstract{
Spatio-temporal data refer to measurements taken across space and time.
In practice, spatio-temporal data can be decomposed into a spatial and
temporal component: at one time, we would select a spatial location and
inspect the temporal trend; at other time, we might select one or
multiple time value(s) and explore the spatial distribution. Ideally, we
could make multiple maps and multiple time series to explore these
together, however, doing all of these actions is complicated when data
arrive fragmented in multiple objects. To make it easy to do all these
tasks, ideally spatial and temporal variables are in a single data
object that we can slice and dice in different ways to conduct different
visualisations. This work suggests a new data structure, \code{cubble},
that uses a nested form and a long form to organise spatial and temporal
variables in a single data object. The new data structure ensures that
data in the two forms are synchronised while been flexible to explore
the two aspect of spatio-temporal data. It also provides capability for
handling data with hierarchical structure, matching data from multiple
sources, constructing interactive graphics, and performing
spatio-temporal transformation. The paper will demonstrate \pkg{cubble}
with examples using Australian climate weather stations, river level
data, and climate reanalysis (ERA5) data.
}

@@

      Nicolas Langrené\\
    BNU-HKBU United International College\\
    2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China\\
  E-mail: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}\\

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time intervals. This type of data can naturally be
considered as a data cube with three axes: spatial identifier, temporal
identifier, and variables. In this cubic formation, space only occupies
one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

\caption[An illustration of different spatio-temporal data layouts]{An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

@@

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} \citep{spacetime} is an attempt for such a
data object, but the spatial and temporal classes it is built on have
now been gradually replaced by more recent implementations. The
\pkg{stars} \citep{stars} package uses an array to wrangle
spatio-temporal data, but the dense array it uses can quickly exhaust
the memory for data in the case of irregular spatial grid. While there
has not yet been a data object that makes it easy to wrangle
spatio-temporal data, there is a strong need for such tools given the
ubiquitous spatial and temporal data collected in the society, i.e.,
climate observations from weather stations, air quality variables from
air monitoring stations. The demand for such tools can also come from
spatial data analysts wishing to incorporate temporal data in their
analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to
overcome the problem with organising variables in different levels when
working with spatio-temporal data. In the package a new data structure,
also called \code{cubble}, is proposed to organise spatial and temporal
variables as two forms of a single data object so that they can be
wrangled separately while synchronised as a whole piece. With this new
data object, analysts can spend less time writing codes to organise
spatial and temporal variables and focus more on the exploratory data
analysis itself. The software is available from the Comprehensive R
Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the main
design and functionality of \pkg{cubble}. Section 3 explains how cubble
deals with more advanced considerations including data with hierarchical
structure, data matching, how cubble fits with existing static and
interactive visualisation tools, and spatio-temporal data
transformation. Section 4 uses Australia weather station data and river
level data as examples to demonstrate the use of \pkg{cubble}. An
example on how \pkg{cubble} handles NetCDF data is also provided.
Section 5 concludes the paper.
